Data Specialization for a Staged Lambda Calculus

### 2015-02-21 paper/implementation.tex
rework section 5

\section {Implementation and Optimization}
\label{sec:implementation}

\begin{concretesyntax}
We have implemented the semantics and splitting algorithm of \lang\ in Standard ML.
In addition to the features shown here, our implementation includes more base types, 
$n$-ary products and sums, if and let statements, and a basic elaboration step.

Applying the rules of \cref{sec:splitting} naively can yield inefficient split programs.
In many cases, the results can be improved by considering special cases in individual rules.
For instance, trivial subterm precomputations needn't be saved at tuples:
\[
\infer {\splittwo {(e_1,e_2)} A {p,l.(r_1,r_2)}}{\splittwo {e_1} A {(),\_.r_1} & \splittwo {e_2} A {p,l.r_2}}
\]
Our implementation uses many optimizations like these.

\end{concretesyntax}

>>>>

\section {Implementation}
\label{sec:implementation}

\begin{abstrsyn}
We have a Standard ML implementation of the \lang\ language, a staged
interpreter, the splitting algorithm, and an interpreter for split programs.
This implementation extends the language described in \ref{sec:semantics} with 
\rmint{}s,
\rmbool{}s,
\texttt{let} statements,
$n$-ary sums and products,
datatype and function declarations, and
deep pattern matching (including $\next{}$ and $\pure{}$ patterns).
The code snippets in this paper are written in our concrete syntax, using 
these additional features when convenient.

We allow staging annotations around declarations. For example,
\begin{lstlisting}
1`atsigngrnd{
  datatype list = Empty | Cons of int * list
  fun part (...) = ...
}
fun qsStaged (...) = ...`
\end{lstlisting}
declares the \texttt{list} datatype and \texttt{part} function at world \bbonep,
by elaborating into:
\begin{lstlisting}
1`val grnd{Empty} = grnd{roll (inj ...)}
val grnd{Cons}  = grnd{fn (...) => roll (inj ...)}
val grnd{part}  = grnd{fn (...) => ...}
val qsStaged    = fn (...) => ...`
\end{lstlisting}

We also have a number of optimizations which drastically improve the readability
of split programs. We split patterns directly, rather than translating them into
lower-level constructs. Finally, we perform many local simplifications during
the splitting algorithm, most notably, not pairing precomputations when one is
known to be $\tup{}$.

%Specifically, our implemention includes parsers, printers, and interpreters

%Its main notable property is that world boundaries are always noted by braces.

%We also emit tuple patterns liberally in split programs.

\end{abstrsyn}
